### 2018-01-21 paper-tourrd3.Rmd
start on paper on custom msgs

---
title: Connecting R with D3 for dynamic graphics, to explore multivariate data with tours
author:
  - name: Michael Kipp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  mkipp271@gmail.com
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
  - name: Ursula Laa
    affiliation: Monash University
    address:
    - Department of Physics
    email:  ursula.laa@monash.edu
bibliography: references.bib
link-citations: true
fig_caption: yes
abstract: >
  The tourr package in R (REF) has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections. It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables. The tourrGui (REF) package provides these sorts of controls, but the interface is programmed with the dated RGtk2 (REF) package. This work explores using custom messages to pass data from R to D3 (REF) for viewing, using the shiny framework.  
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

```{r cache=FALSE, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
#packages:
library(bookdown)
library(knitcitations)
library(RefManageR)
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, kfigr.link=TRUE, kfigr.prefix=TRUE, cache=TRUE, fig.env = TRUE, fig.cap=TRUE, fig.height=3)

options("citation_format" = "pandoc")

BibOptions(check.entries = FALSE, style = "markdown", bib.style = "alphabetic", cite.style = 'alphabetic')
```

## Introduction

The tour algorithm [@gt_pp;@gt_pp_mc] is a way of systematically generating and displaying projections of high-dimensional spaces in order for the viewer to examine the multivariate distribution of data. It can do this either randomly, or by picking projections judged interesting according to some criterion or index function. The tourr package [@tourr] provides the computing and display in R [@R,@ihaka:1996] to make several types of tours, grand, guided, little and local. The projection dimension can be chosen between one and the number of variables in the data. The display, though, has no capacity for interaction. The viewer can watch the tour like a movie, but not pause it and restart, or change tour type, or number of variables.

These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2]. This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like shiny [@shiny]. To display dynamic graphics though, still takes some work. This work explores the use of D3 [@2011-d3] as the display engine.

## Section title in sentence case

This section may contain a figure such as Figure \ref{figure:rlogo}.

\begin{figure}[htbp]
  \centering
  \caption{The logo of R.}
  \label{figure:rlogo}
\end{figure}

## Another section

There will likely be several sections, perhaps including code snippets, such as:

```{r}
x <- 1:10
x
```

## Summary

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

## References

<div id="refs"></div>

### 2018-01-21 paper-tourrd3.tex
start on paper on custom msgs

% !TeX root = RJwrapper.tex
\title{Connecting R with D3 for dynamic graphics, to explore multivariate data
with tours}
\author{by Michael Kipp, Dianne Cook, Ursula Laa}

\maketitle

\abstract{%
The tourr package in R (REF) has several algorithms and displays for
showing multivariate data as a sequence of low-dimensional projections.
It can display as a movie but has no capacity for interaction, such as
stop/go, change tour type, drop/add variables. The tourrGui (REF)
package provides these sorts of controls, but the interface is
programmed with the dated RGtk2 (REF) package. This work explores using
custom messages to pass data from R to D3 (REF) for viewing, using the
shiny framework.
}

\subsection{Introduction}\label{introduction}

The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995};
Cook et al. \protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. \protect\hyperlink{ref-tourr}{2011})
provides the computing and display in R (R Core Team
\protect\hyperlink{ref-R}{2012},Ihaka and Gentleman
(\protect\hyperlink{ref-ihaka:1996}{1996})) to make several types of
tours, grand, guided, little and local. The projection dimension can be
chosen between one and the number of variables in the data. The display,
though, has no capacity for interaction. The viewer can watch the tour
like a movie, but not pause it and restart, or change tour type, or
number of variables.

These interactive controls were provided with the tourrGui package
(Huang, Cook, and Wickham \protect\hyperlink{ref-tourrGui}{2012}), with
was programmed with the RGtk2 package (Lawrence and Temple Lang
\protect\hyperlink{ref-RGtk2}{2010}). This is not the toolkit of choice
today, and has been superceded with primarily web-capable tools, like
shiny (Chang et al. \protect\hyperlink{ref-shiny}{2017}). To display
dynamic graphics though, still takes some work. This work explores the
use of D3 (Heer \protect\hyperlink{ref-2011-d3}{2011}) as the display
engine.

\subsection{Section title in sentence
case}\label{section-title-in-sentence-case}

This section may contain a figure such as Figure \ref{figure:rlogo}.

\begin{figure}[htbp]
  \centering
  \caption{The logo of R.}
  \label{figure:rlogo}
\end{figure}

\subsection{Another section}\label{another-section}

There will likely be several sections, perhaps including code snippets,
such as:

\begin{Schunk}
\begin{Soutput}
#>  [1]  1  2  3  4  5  6  7  8  9 10
\end{Soutput}
\end{Schunk}

\subsection{Summary}\label{summary}

This file is only a basic article template. For full details of
\emph{The R Journal} style and information on how to prepare your
article for submission, see the
\href{https://journal.r-project.org/share/author-guide.pdf}{Instructions
for Authors}.

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-shiny}{}
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
McPherson. 2017. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

\hypertarget{ref-gt_pp}{}
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (4): 155--72.

\hypertarget{ref-gt_pp_mc}{}
Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''

\hypertarget{ref-2011-d3}{}
Heer, Michael Bostock AND Vadim Ogievetsky AND Jeffrey. 2011. ``D3:
Data-Driven Documents.'' \emph{IEEE Trans. Visualization \& Comp.
Graphics (Proc. InfoVis)}. \url{http://vis.stanford.edu/papers/d3}.

\hypertarget{ref-tourrGui}{}
Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
gWidgets Gui for the Tour to Explore High-Dimensional Data Using
Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
(6): 1--12.

\hypertarget{ref-ihaka:1996}{}
Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
Analysis and Graphics.'' \emph{Journal of Computational and Graphical
Statistics} 5 (3): 299--314.

\hypertarget{ref-RGtk2}{}
Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
37 (8): 1--52. \url{http://www.jstatsoft.org/v37/i08/}.

\hypertarget{ref-R}{}
R Core Team. 2012. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

\hypertarget{ref-tourr}{}
Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``Tourr: An R Package for Exploring Multivariate Data with
Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.

\address{%
Michael Kipp\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:mkipp271@gmail.com}{\nolinkurl{mkipp271@gmail.com}}

\address{%
Dianne Cook\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}

\address{%
Ursula Laa\\
Monash University\\
Department of Physics\\
}
\href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}

### 2018-02-16 paper-tourrd3.Rmd
finished?

@@
 abstract: >
-  The tourr package in R (REF) has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections.
+  The tourr package in R has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections.
 It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables.
-The tourrGui (REF) package provides these sorts of controls, but the interface is programmed with the dated RGtk2 (REF) package.
-This work explores using custom messages to pass data from R to D3 (REF) for viewing, using the shiny framework.
+The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package.
+This work explores using custom messages to pass data from R to D3  for viewing, using the shiny framework.
 preamble: >
@@
 
+```{r}
+library(shiny)
+library(tourr)
+```
+
 ## Introduction
 
-The tour algorithm [@gt_pp;@gt_pp_mc] is a way of systematically generating and displaying projections of high-dimensional spaces in order for the viewer to examine the multivariate distribution of data.
+The tour algorithm [@gt_pp; @gt_pp_mc] is a way of systematically generating and displaying projections of high-dimensional spaces in order for the viewer to examine the multivariate distribution of data.
 It can do this either randomly, or by picking projections judged interesting according to some criterion or index function.
-The tourr package [@tourr] provides the computing and display in R [@R,@ihaka:1996] to make several types of tours, grand, guided, little and local.
+The tourr package [@tourr] provides the computing and display in R [@R; @ihaka:1996] to make several types of tours, grand, guided, little and local.
 The projection dimension can be chosen between one and the number of variables in the data.
@@
 This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like shiny [@shiny].
-To display dynamic graphics though, still takes some work.
-This work explores the use of D3 [@2011-d3] as the display engine.
+To display dynamic graphics though, is not straight-forward.
+This paper explains how to use D3 [@D3] as the display engine, in a shiny GUI, using custom message passing between server and client.
 
-## Section title in sentence case
+## Creating a tour, with the tourr package
 
-This section may contain a figure such as Figure \ref{figure:rlogo}.
+The \pkg{tourr} package [@tourr] is an R implementation of the tour algorithms discussed in @gt_pp_mc.
+It includes methods for geodesic interpolation and basis generation, as well as an implementation of the simulated annealing algorithm to optimise projection pursuit indices for the guided tour.
+The tour can be displayed directly in the R graphics device, for example:
 
-\begin{figure}[htbp]
-  \centering
-  \caption{The logo of R.}
-  \label{figure:rlogo}
+```{r eval=FALSE, echo=TRUE}
+# quartz() # to display on a Mac
+# X11() # For windows
+# The Rstudio graphics device cannot be used
+animate_dist(flea[, 1:6], center = FALSE)
+```
+
+generating the animation in Figure \ref{tour}.
+
+\begin{figure}[ht]
+EMBED MOVIE HERE
 \end{figure}
 
-## Another section
+A tour path is a smooth sequence of projection matrices, $p\times d$, that when combined with data, $n\times p$, and a rendering method, produces a steady stream of $d$-dimensional views of the data.
+Each tour is initialised with the `new_tour()` method, which instantiates a tour object and takes as arguments the data $X$, the tour method, e.g.
+`guided_tour()`, and the starting basis.
+Once initialised, a new target plane is chosen, and a series of steps along a geodesic path from starting to target plane are generated by interpolation.
 
-There will likely be several sections, perhaps including code snippets, such as:
+This requires a series of calls to the tour object producing the series of projections.
+The steps are discrete, of size given by $\omega/\Delta$, where $\omega$ denotes the angular velocity of the geodesic interpolation, and $\Delta$ is a parameter denoting frames per second, reflecting the rendering speed of the device in use.
+The $\Delta$ parameter can be thought of as the frames per second, while $\omega$ affects the speed at which the tour moves through the projection space.
+For our purposes, $\Delta$, `fps` in the code, is set at 25, while the $\omega$ can be adjusted by the user.
 
-```{r}
-x <- 1:10
-x
+## Connecting the tour projections to D3 display using `sendCustomMessage`
+
+D3.js (Data-Driven Documents) (REF) is a JavaScript library for manipulating documents based on data.
+The advantages of D3 are similar to those provided by Shiny: namely, an industry standard with rich array of powerful, easy to use methods and widgets that can be displayed on a wide variety of devices, with a large user base.
+D3 works on data objects in the JavaScript Object Notation (JSON) format, which are then parsed and used to display customisable data visualisations.
+
+The new implementation of the tour uses D3 to render each projection step returned by R, focusing on 2D projections as a test case.
+It does this by drawing and re-drawing a scatterplot with dots (or circles in D3 language) and providing SVG objects for the web browser to render.
+
+There two functions provided by the Shiny framework to transport data be- tween R and javascript: `session$sendCustomMessage()` in R, and the corresponding `Shiny.addCustomMessageHandler()` in Javascript.
+Whenever the former is executed in R, the latter function will execute a code block in JS.
+There are many examples of such functions being used to pass arbitrary data from an R app to a JS front-end, few examples exist of this basic functionality to update a D3 animation in real-time.
+
+The data format expected by D3 is in JSON format, which combines two basic program- ming paradigms: a collection of name/value pairs, and an ordered list of values.
+R's preferred data formats include data frames, vectors and matrices.
+Every time a new projection has been calculated with the tour path, the resulting matrix needs to be converted to JSON and sent to D3.
+The code to send the D3 data looks like this:
+
 ```
+session$sendCustomMessage(type = "data", message = toJSON(j))
+```
+
+This code is from the observe environment from the `server.R` file.
+It converts the projection matrix to JSON format, and sends it to JavaScript with the id data.
+When parsed in D3 by its `data()` method, it is converted back into a logical 2D array where the columns are queried first, then the rows.
+If column names are included in the JSON, the column indices are strings; otherwise they are integers starting from 1.
+All of the code required to render the scatterplots and legends, along with colours, is JavaScript code in the file `d3anim.js`. In particular, the data from R is handled with the following code:
+
+```
+Shiny.addCustomMessageHandler("data",
+    function(message) {
+        /* D3 scatterplot is drawn and re-drawn using the
+            data sent from the server. */
+}
+```
+
+Every time the message is sent (25 times per second), the code-block is run.
+
+## Getting projections
+
+The `observeEvent` Shiny method defines a code block to be run whenever some input value changes. The following code snippet restarts a tour using a random basis:
+
+```
+observeEvent(input$restart_random,
+  {
+    p <- length(input$variables)
+    b <- matrix(runif(2*p), p, 2)
+    rv$tour <- 
+      new_tour(as.matrix(rv$d[input$variables]),
+              choose_tour(input$type, 
+              input$guidedIndex,
+              c(rv$class[[1]]), 
+              input$scagType), b)
+})
+```
+
+The projections are calculated using the tour object in an `observe()` environment, which repeatedly runs the code until a reactive variable is changed, at which point it is is "invalidated" and then re-started.
+The projections are calculated using the following code
+block:
+
+```
+observe({
+  tour <- rv$tour
+  aps <- rv$aps
+  step <- tour(aps / fps)
+  if (!is.null(step)) {
+    invalidateLater(1000 / fps)
+    j <- center(rv$mat %*% step$proj)
+    j <- cbind(j, class = rv$class)
+    colnames(j) <- NULL
+    session$sendCustomMessage(type = "data", 
+      message = toJSON(j))
+  }
+  else {
+    if (length(rv$mat[1, ]) < 3) {
+    session$sendCustomMessage(type = "debug", 
+      message = "Error: Need > 2 variables.")
+    } else {
+    session$sendCustomMessage(type = "debug", 
+    message = "Guided tour finished: no better bases found.")
+    }
+  }
+})
+```
+
+## Pros and cons
+
+The D3 canvas makes for smooth drawing and re-drawing of the data projections.
+Adding a graphical user interface (GUI) around the display is straightforward with the shiny package, e.g.
+control elements such as stop/go, increase/decrease speed, change tour type, add/remove variables from the mix.
+
+The main disadvantage is that the speed is inconsistent, as server and client play tag to keep up with each other, and the display cannot handle many observations.
+Noticeable slow down was oberved with 2000 points.
+
+## Code
+
+The code is available at [https://github.com/makipp/ TourrGuiD3], and the source material for this paper is available at [https://github.com/dicook/paper-tourrd3].
 
@@
 
-This file is only a basic article template.
-For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).
+The `sendCustomMessage` function provides a way to share a tour path with the D3 renderer, and embed it in a shiny GUI providing controls such as top/go, increase/decrease speed, change tour type, add/remove variables.
+However, the approach can result in inconsistent speed which appears as jerkiness in the motion, and it isn't suitable for large numbers of observations.
 

### 2018-02-16 paper-tourrd3.tex
finished?

@@
 \abstract{%
-The tourr package in R (REF) has several algorithms and displays for
-showing multivariate data as a sequence of low-dimensional projections.
-It can display as a movie but has no capacity for interaction, such as
-stop/go, change tour type, drop/add variables. The tourrGui (REF)
-package provides these sorts of controls, but the interface is
-programmed with the dated RGtk2 (REF) package. This work explores using
-custom messages to pass data from R to D3 (REF) for viewing, using the
-shiny framework.
+The tourr package in R has several algorithms and displays for showing
+multivariate data as a sequence of low-dimensional projections. It can
+display as a movie but has no capacity for interaction, such as stop/go,
+change tour type, drop/add variables. The tourrGui package provides
+these sorts of controls, but the interface is programmed with the dated
+RGtk2 package. This work explores using custom messages to pass data
+from R to D3 for viewing, using the shiny framework.
 }
@@
 provides the computing and display in R (R Core Team
-\protect\hyperlink{ref-R}{2012},Ihaka and Gentleman
-(\protect\hyperlink{ref-ihaka:1996}{1996})) to make several types of
+\protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
+\protect\hyperlink{ref-ihaka:1996}{1996}) to make several types of
 tours, grand, guided, little and local. The projection dimension can be
@@
 shiny (Chang et al. \protect\hyperlink{ref-shiny}{2017}). To display
-dynamic graphics though, still takes some work. This work explores the
-use of D3 (Heer \protect\hyperlink{ref-2011-d3}{2011}) as the display
-engine.
+dynamic graphics though, is not straight-forward. This paper explains
+how to use D3 (Bostock, Ogievetsky, and Heer
+\protect\hyperlink{ref-D3}{2011}) as the display engine, in a shiny GUI,
+using custom message passing between server and client.
 
-\subsection{Section title in sentence
-case}\label{section-title-in-sentence-case}
+\subsection{Creating a tour, with the tourr
+package}\label{creating-a-tour-with-the-tourr-package}
 
-This section may contain a figure such as Figure \ref{figure:rlogo}.
+The \pkg{tourr} package (Wickham et al.
+\protect\hyperlink{ref-tourr}{2011}) is an R implementation of the tour
+algorithms discussed in Cook et al.
+(\protect\hyperlink{ref-gt_pp_mc}{2007}). It includes methods for
+geodesic interpolation and basis generation, as well as an
+implementation of the simulated annealing algorithm to optimise
+projection pursuit indices for the guided tour. The tour can be
+displayed directly in the R graphics device, for example:
 
-\begin{figure}[htbp]
-  \centering
-  \caption{The logo of R.}
-  \label{figure:rlogo}
+\begin{Schunk}
+\begin{Sinput}
+# quartz() # to display on a Mac
+# X11() # For windows
+# The Rstudio graphics device cannot be used
+animate_dist(flea[, 1:6], center = FALSE)
+\end{Sinput}
+\end{Schunk}
+
+generating the animation in Figure \ref{tour}.
+
+\begin{figure}[ht]
+EMBED MOVIE HERE
 \end{figure}
 
-\subsection{Another section}\label{another-section}
+A tour path is a smooth sequence of projection matrices, \(p\times d\),
+that when combined with data, \(n\times p\), and a rendering method,
+produces a steady stream of \(d\)-dimensional views of the data. Each
+tour is initialised with the \texttt{new\_tour()} method, which
+instantiates a tour object and takes as arguments the data \(X\), the
+tour method, e.g. \texttt{guided\_tour()}, and the starting basis. Once
+initialised, a new target plane is chosen, and a series of steps along a
+geodesic path from starting to target plane are generated by
+interpolation.
 
-There will likely be several sections, perhaps including code snippets,
-such as:
+This requires a series of calls to the tour object producing the series
+of projections. The steps are discrete, of size given by
+\(\omega/\Delta\), where \(\omega\) denotes the angular velocity of the
+geodesic interpolation, and \(\Delta\) is a parameter denoting frames
+per second, reflecting the rendering speed of the device in use. The
+\(\Delta\) parameter can be thought of as the frames per second, while
+\(\omega\) affects the speed at which the tour moves through the
+projection space. For our purposes, \(\Delta\), \texttt{fps} in the
+code, is set at 25, while the \(\omega\) can be adjusted by the user.
 
-\begin{Schunk}
-\begin{Soutput}
-#>  [1]  1  2  3  4  5  6  7  8  9 10
-\end{Soutput}
-\end{Schunk}
+\subsection{\texorpdfstring{Connecting the tour projections to D3
+display using
+\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}
+
+D3.js (Data-Driven Documents) (REF) is a JavaScript library for
+manipulating documents based on data. The advantages of D3 are similar
+to those provided by Shiny: namely, an industry standard with rich array
+of powerful, easy to use methods and widgets that can be displayed on a
+wide variety of devices, with a large user base. D3 works on data
+objects in the JavaScript Object Notation (JSON) format, which are then
+parsed and used to display customisable data visualisations.
+
+The new implementation of the tour uses D3 to render each projection
+step returned by R, focusing on 2D projections as a test case. It does
+this by drawing and re-drawing a scatterplot with dots (or circles in D3
+language) and providing SVG objects for the web browser to render.
+
+There two functions provided by the Shiny framework to transport data
+be- tween R and javascript: \texttt{session\$sendCustomMessage()} in R,
+and the corresponding \texttt{Shiny.addCustomMessageHandler()} in
+Javascript. Whenever the former is executed in R, the latter function
+will execute a code block in JS. There are many examples of such
+functions being used to pass arbitrary data from an R app to a JS
+front-end, few examples exist of this basic functionality to update a D3
+animation in real-time.
+
+The data format expected by D3 is in JSON format, which combines two
+basic program- ming paradigms: a collection of name/value pairs, and an
+ordered list of values. R's preferred data formats include data frames,
+vectors and matrices. Every time a new projection has been calculated
+with the tour path, the resulting matrix needs to be converted to JSON
+and sent to D3. The code to send the D3 data looks like this:
+
+\begin{verbatim}
+session$sendCustomMessage(type = "data", message = toJSON(j))
+\end{verbatim}
+
+This code is from the observe environment from the \texttt{server.R}
+file. It converts the projection matrix to JSON format, and sends it to
+JavaScript with the id data. When parsed in D3 by its \texttt{data()}
+method, it is converted back into a logical 2D array where the columns
+are queried first, then the rows. If column names are included in the
+JSON, the column indices are strings; otherwise they are integers
+starting from 1. All of the code required to render the scatterplots and
+legends, along with colours, is JavaScript code in the file
+\texttt{d3anim.js}. In particular, the data from R is handled with the
+following code:
+
+\begin{verbatim}
+Shiny.addCustomMessageHandler("data",
+    function(message) {
+        /* D3 scatterplot is drawn and re-drawn using the
+            data sent from the server. */
+}
+\end{verbatim}
+
+Every time the message is sent (25 times per second), the code-block is
+run.
+
+\subsection{Getting projections}\label{getting-projections}
+
+The \texttt{observeEvent} Shiny method defines a code block to be run
+whenever some input value changes. The following code snippet restarts a
+tour using a random basis:
+
+\begin{verbatim}
+observeEvent(input$restart_random,
+  {
+    p <- length(input$variables)
+    b <- matrix(runif(2*p), p, 2)
+    rv$tour <- 
+      new_tour(as.matrix(rv$d[input$variables]),
+              choose_tour(input$type, 
+              input$guidedIndex,
+              c(rv$class[[1]]), 
+              input$scagType), b)
+})
+\end{verbatim}
+
+The projections are calculated using the tour object in an
+\texttt{observe()} environment, which repeatedly runs the code until a
+reactive variable is changed, at which point it is is ``invalidated''
+and then re-started. The projections are calculated using the following
+code block:
+
+\begin{verbatim}
+observe({
+  tour <- rv$tour
+  aps <- rv$aps
+  step <- tour(aps / fps)
+  if (!is.null(step)) {
+    invalidateLater(1000 / fps)
+    j <- center(rv$mat %*% step$proj)
+    j <- cbind(j, class = rv$class)
+    colnames(j) <- NULL
+    session$sendCustomMessage(type = "data", 
+      message = toJSON(j))
+  }
+  else {
+    if (length(rv$mat[1, ]) < 3) {
+    session$sendCustomMessage(type = "debug", 
+      message = "Error: Need > 2 variables.")
+    } else {
+    session$sendCustomMessage(type = "debug", 
+    message = "Guided tour finished: no better bases found.")
+    }
+  }
+})
+\end{verbatim}
+
+\subsection{Pros and cons}\label{pros-and-cons}
+
+The D3 canvas makes for smooth drawing and re-drawing of the data
+projections. Adding a graphical user interface (GUI) around the display
+is straightforward with the shiny package, e.g.~control elements such as
+stop/go, increase/decrease speed, change tour type, add/remove variables
+from the mix.
+
+The main disadvantage is that the speed is inconsistent, as server and
+client play tag to keep up with each other, and the display cannot
+handle many observations. Noticeable slow down was oberved with 2000
+points.
+
+\subsection{Code}\label{code}
+
+The code is available at {[}\url{https://github.com/makipp/}
+TourrGuiD3{]}, and the source material for this paper is available at
+{[}\url{https://github.com/dicook/paper-tourrd3}{]}.
 
@@
 
-This file is only a basic article template. For full details of
-\emph{The R Journal} style and information on how to prepare your
-article for submission, see the
-\href{https://journal.r-project.org/share/author-guide.pdf}{Instructions
-for Authors}.
+The \texttt{sendCustomMessage} function provides a way to share a tour
+path with the D3 renderer, and embed it in a shiny GUI providing
+controls such as top/go, increase/decrease speed, change tour type,
+add/remove variables. However, the approach can result in inconsistent
+speed which appears as jerkiness in the motion, and it isn't suitable
+for large numbers of observations.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-D3}{}
+Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
+Data-Driven Documents.'' \emph{IEEE Transactions on Visualization and
+Computer Graphics} 17 (12): 2301--9.
+\url{http://vis.stanford.edu/papers/d3}.
+
 \hypertarget{ref-shiny}{}
@@
 ``Grand Tours, Projection Pursuit Guided Tours and Manual Controls.''
-
-\hypertarget{ref-2011-d3}{}
-Heer, Michael Bostock AND Vadim Ogievetsky AND Jeffrey. 2011. ``D3:
-Data-Driven Documents.'' \emph{IEEE Trans. Visualization \& Comp.
-Graphics (Proc. InfoVis)}. \url{http://vis.stanford.edu/papers/d3}.
 
@@
 \hypertarget{ref-R}{}
-R Core Team. 2012. \emph{R: A Language and Environment for Statistical
+R Core Team. 2018. \emph{R: A Language and Environment for Statistical
 Computing}. Vienna, Austria: R Foundation for Statistical Computing.

### 2018-02-19 paper-tourrd3.Rmd
Small changes/fixes in the text.

@@
 The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package.
-This work explores using custom messages to pass data from R to D3  for viewing, using the shiny framework.
+This work explores using custom messages to pass data from R to D3  for viewing, using the Shiny framework.
 preamble: >
@@
 These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2].
-This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like shiny [@shiny].
+This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like Shiny [@shiny].
 To display dynamic graphics though, is not straight-forward.
-This paper explains how to use D3 [@D3] as the display engine, in a shiny GUI, using custom message passing between server and client.
+This paper explains how to use D3 [@D3] as the display engine, in a Shiny GUI, using custom message passing between server and client.
 
@@
 
-The new implementation of the tour uses D3 to render each projection step returned by R, focusing on 2D projections as a test case.
+The new implementation of the tour interface uses D3 to render each projection step returned by R, focusing on 2D projections as a test case.
 It does this by drawing and re-drawing a scatterplot with dots (or circles in D3 language) and providing SVG objects for the web browser to render.
 
-There two functions provided by the Shiny framework to transport data be- tween R and javascript: `session$sendCustomMessage()` in R, and the corresponding `Shiny.addCustomMessageHandler()` in Javascript.
+There are two functions provided by the Shiny framework to transport data between R and JavaScript: `session$sendCustomMessage()` in R, and the corresponding `Shiny.addCustomMessageHandler()` in JavaScript.
 Whenever the former is executed in R, the latter function will execute a code block in JS.
@@
 
-The data format expected by D3 is in JSON format, which combines two basic program- ming paradigms: a collection of name/value pairs, and an ordered list of values.
+The data format expected by D3 is in JSON format, which combines two basic programming paradigms: a collection of name/value pairs, and an ordered list of values.
 R's preferred data formats include data frames, vectors and matrices.
@@
 This code is from the observe environment from the `server.R` file.
-It converts the projection matrix to JSON format, and sends it to JavaScript with the id data.
+It converts the matrix of projected data points to JSON format, and sends it to JavaScript with the id data.
 When parsed in D3 by its `data()` method, it is converted back into a logical 2D array where the columns are queried first, then the rows.
@@
 The D3 canvas makes for smooth drawing and re-drawing of the data projections.
-Adding a graphical user interface (GUI) around the display is straightforward with the shiny package, e.g.
+Adding a graphical user interface (GUI) around the display is straightforward with the Shiny package, e.g.
 control elements such as stop/go, increase/decrease speed, change tour type, add/remove variables from the mix.
@@
 The main disadvantage is that the speed is inconsistent, as server and client play tag to keep up with each other, and the display cannot handle many observations.
-Noticeable slow down was oberved with 2000 points.
+Noticeable slow down was oberved with 2000 points, the main reason being the rendering time required for the large number of SVG circle elements.
+The situation can be improved when using a single HTML5 canvas element to draw the scatter points, significantly reducing the rendering time.
 
@@
 
-The `sendCustomMessage` function provides a way to share a tour path with the D3 renderer, and embed it in a shiny GUI providing controls such as top/go, increase/decrease speed, change tour type, add/remove variables.
+The `sendCustomMessage` function provides a way to share a tour path with the D3 renderer, and embed it in a Shiny GUI providing controls such as stop/go, increase/decrease speed, change tour type, add/remove variables.
 However, the approach can result in inconsistent speed which appears as jerkiness in the motion, and it isn't suitable for large numbers of observations.

### 2018-02-19 paper-tourrd3.tex
Small changes/fixes in the text.

@@
 RGtk2 package. This work explores using custom messages to pass data
-from R to D3 for viewing, using the shiny framework.
+from R to D3 for viewing, using the Shiny framework.
 }
@@
 
-The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995};
-Cook et al. \protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of
+The tour algorithm (Cook et al. 1995; Cook et al. 2007) is a way of
 systematically generating and displaying projections of high-dimensional
@@
 judged interesting according to some criterion or index function. The
-tourr package (Wickham et al. \protect\hyperlink{ref-tourr}{2011})
-provides the computing and display in R (R Core Team
-\protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
-\protect\hyperlink{ref-ihaka:1996}{1996}) to make several types of
-tours, grand, guided, little and local. The projection dimension can be
-chosen between one and the number of variables in the data. The display,
-though, has no capacity for interaction. The viewer can watch the tour
-like a movie, but not pause it and restart, or change tour type, or
-number of variables.
+tourr package (Wickham et al. 2011) provides the computing and display
+in R (R Core Team 2018; Ihaka and Gentleman 1996) to make several types
+of tours, grand, guided, little and local. The projection dimension can
+be chosen between one and the number of variables in the data. The
+display, though, has no capacity for interaction. The viewer can watch
+the tour like a movie, but not pause it and restart, or change tour
+type, or number of variables.
 
 These interactive controls were provided with the tourrGui package
-(Huang, Cook, and Wickham \protect\hyperlink{ref-tourrGui}{2012}), with
-was programmed with the RGtk2 package (Lawrence and Temple Lang
-\protect\hyperlink{ref-RGtk2}{2010}). This is not the toolkit of choice
-today, and has been superceded with primarily web-capable tools, like
-shiny (Chang et al. \protect\hyperlink{ref-shiny}{2017}). To display
-dynamic graphics though, is not straight-forward. This paper explains
-how to use D3 (Bostock, Ogievetsky, and Heer
-\protect\hyperlink{ref-D3}{2011}) as the display engine, in a shiny GUI,
-using custom message passing between server and client.
+(Huang, Cook, and Wickham 2012), with was programmed with the RGtk2
+package (Lawrence and {Temple Lang} 2010). This is not the toolkit of
+choice today, and has been superceded with primarily web-capable tools,
+like Shiny (Chang et al. 2017). To display dynamic graphics though, is
+not straight-forward. This paper explains how to use D3 (Bostock,
+Ogievetsky, and Heer 2011) as the display engine, in a Shiny GUI, using
+custom message passing between server and client.
 
@@
 
-The \pkg{tourr} package (Wickham et al.
-\protect\hyperlink{ref-tourr}{2011}) is an R implementation of the tour
-algorithms discussed in Cook et al.
-(\protect\hyperlink{ref-gt_pp_mc}{2007}). It includes methods for
-geodesic interpolation and basis generation, as well as an
+The \pkg{tourr} package (Wickham et al. 2011) is an R implementation of
+the tour algorithms discussed in Cook et al. (2007). It includes methods
+for geodesic interpolation and basis generation, as well as an
 implementation of the simulated annealing algorithm to optimise
@@
 
-\subsection{\texorpdfstring{Connecting the tour projections to D3
-display using
-\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}
+\subsection{Connecting the tour projections to D3 display using
+\texttt{sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}
 
@@
 
-The new implementation of the tour uses D3 to render each projection
-step returned by R, focusing on 2D projections as a test case. It does
-this by drawing and re-drawing a scatterplot with dots (or circles in D3
-language) and providing SVG objects for the web browser to render.
+The new implementation of the tour interface uses D3 to render each
+projection step returned by R, focusing on 2D projections as a test
+case. It does this by drawing and re-drawing a scatterplot with dots (or
+circles in D3 language) and providing SVG objects for the web browser to
+render.
 
-There two functions provided by the Shiny framework to transport data
-be- tween R and javascript: \texttt{session\$sendCustomMessage()} in R,
-and the corresponding \texttt{Shiny.addCustomMessageHandler()} in
-Javascript. Whenever the former is executed in R, the latter function
+There are two functions provided by the Shiny framework to transport
+data between R and JavaScript: \texttt{session\$sendCustomMessage()} in
+R, and the corresponding \texttt{Shiny.addCustomMessageHandler()} in
+JavaScript. Whenever the former is executed in R, the latter function
 will execute a code block in JS. There are many examples of such
@@
 The data format expected by D3 is in JSON format, which combines two
-basic program- ming paradigms: a collection of name/value pairs, and an
+basic programming paradigms: a collection of name/value pairs, and an
 ordered list of values. R's preferred data formats include data frames,
@@
 This code is from the observe environment from the \texttt{server.R}
-file. It converts the projection matrix to JSON format, and sends it to
-JavaScript with the id data. When parsed in D3 by its \texttt{data()}
-method, it is converted back into a logical 2D array where the columns
-are queried first, then the rows. If column names are included in the
-JSON, the column indices are strings; otherwise they are integers
-starting from 1. All of the code required to render the scatterplots and
-legends, along with colours, is JavaScript code in the file
-\texttt{d3anim.js}. In particular, the data from R is handled with the
-following code:
+file. It converts the matrix of projected data points to JSON format,
+and sends it to JavaScript with the id data. When parsed in D3 by its
+\texttt{data()} method, it is converted back into a logical 2D array
+where the columns are queried first, then the rows. If column names are
+included in the JSON, the column indices are strings; otherwise they are
+integers starting from 1. All of the code required to render the
+scatterplots and legends, along with colours, is JavaScript code in the
+file \texttt{d3anim.js}. In particular, the data from R is handled with
+the following code:
 
@@
 projections. Adding a graphical user interface (GUI) around the display
-is straightforward with the shiny package, e.g.~control elements such as
+is straightforward with the Shiny package, e.g.~control elements such as
 stop/go, increase/decrease speed, change tour type, add/remove variables
@@
 handle many observations. Noticeable slow down was oberved with 2000
-points.
+points, the main reason being the rendering time required for the large
+number of SVG circle elements. The situation can be improved when using
+a single HTML5 canvas element to draw the scatter points, significantly
+reducing the rendering time.
 
@@
 The \texttt{sendCustomMessage} function provides a way to share a tour
-path with the D3 renderer, and embed it in a shiny GUI providing
-controls such as top/go, increase/decrease speed, change tour type,
+path with the D3 renderer, and embed it in a Shiny GUI providing
+controls such as stop/go, increase/decrease speed, change tour type,
 add/remove variables. However, the approach can result in inconsistent
@@
 
-\hypertarget{refs}{}
-\hypertarget{ref-D3}{}
+\hyperdef{}{refs}{}
+
 Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
@@
 
-\hypertarget{ref-shiny}{}
 Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
@@
 
-\hypertarget{ref-gt_pp}{}
 Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
@@
 
-\hypertarget{ref-gt_pp_mc}{}
 Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
@@
 
-\hypertarget{ref-tourrGui}{}
 Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
-gWidgets Gui for the Tour to Explore High-Dimensional Data Using
+GWidgets GUI for the Tour to Explore High-Dimensional Data Using
 Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
@@
 
-\hypertarget{ref-ihaka:1996}{}
 Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
@@
 
-\hypertarget{ref-RGtk2}{}
-Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
+Lawrence, Michael, and Duncan {Temple Lang}. 2010. ``RGtk2: A Graphical
 User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
@@
 
-\hypertarget{ref-R}{}
 R Core Team. 2018. \emph{R: A Language and Environment for Statistical
@@
 
-\hypertarget{ref-tourr}{}
 Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.

### 2018-03-02 paper-tourrd3.Rmd
Small changes only

@@
     address:
-    - Department of Physics
+    - School of Physics and Astronomy
     email:  ursula.laa@monash.edu
@@
 It can do this either randomly, or by picking projections judged interesting according to some criterion or index function.
-The tourr package [@tourr] provides the computing and display in R [@R; @ihaka:1996] to make several types of tours, grand, guided, little and local.
+The tourr package [@tourr] provides the computing and display in R [@R; @ihaka:1996] to make several types of tours: grand, guided, little and local.
 The projection dimension can be chosen between one and the number of variables in the data.
@@
 To display dynamic graphics though, is not straight-forward.
-This paper explains how to use D3 [@D3] as the display engine, in a Shiny GUI, using custom message passing between server and client.
+This paper explains how to use D3 [@D3] as the display engine in a Shiny GUI, using custom message passing between server and client.
 
@@
 
-A tour path is a smooth sequence of projection matrices, $p\times d$, that when combined with data, $n\times p$, and a rendering method, produces a steady stream of $d$-dimensional views of the data.
+A tour path is a smooth sequence of projection matrices, $p\times d$, that when combined with a matrix of n data points, $n\times p$, and a rendering method, produces a steady stream of $d$-dimensional views of the data.
 Each tour is initialised with the `new_tour()` method, which instantiates a tour object and takes as arguments the data $X$, the tour method, e.g.
@@
 
-D3.js (Data-Driven Documents) (REF) is a JavaScript library for manipulating documents based on data.
+D3.js (Data-Driven Documents) [@D3] is a JavaScript library for manipulating documents based on data.
 The advantages of D3 are similar to those provided by Shiny: namely, an industry standard with rich array of powerful, easy to use methods and widgets that can be displayed on a wide variety of devices, with a large user base.
@@
 
-The projections are calculated using the tour object in an `observe()` environment, which repeatedly runs the code until a reactive variable is changed, at which point it is is "invalidated" and then re-started.
+The projections are calculated using the tour object in an `observe()` environment, which repeatedly runs the code until a reactive variable is changed, at which point it is "invalidated" and then re-started.
 The projections are calculated using the following code

### 2018-03-02 paper-tourrd3.tex
Small changes only

@@
 in R (R Core Team 2018; Ihaka and Gentleman 1996) to make several types
-of tours, grand, guided, little and local. The projection dimension can
+of tours: grand, guided, little and local. The projection dimension can
 be chosen between one and the number of variables in the data. The
@@
 not straight-forward. This paper explains how to use D3 (Bostock,
-Ogievetsky, and Heer 2011) as the display engine, in a Shiny GUI, using
+Ogievetsky, and Heer 2011) as the display engine in a Shiny GUI, using
 custom message passing between server and client.
@@
 A tour path is a smooth sequence of projection matrices, \(p\times d\),
-that when combined with data, \(n\times p\), and a rendering method,
-produces a steady stream of \(d\)-dimensional views of the data. Each
-tour is initialised with the \texttt{new\_tour()} method, which
-instantiates a tour object and takes as arguments the data \(X\), the
-tour method, e.g. \texttt{guided\_tour()}, and the starting basis. Once
-initialised, a new target plane is chosen, and a series of steps along a
-geodesic path from starting to target plane are generated by
+that when combined with a matrix of n data points, \(n\times p\), and a
+rendering method, produces a steady stream of \(d\)-dimensional views of
+the data. Each tour is initialised with the \texttt{new\_tour()} method,
+which instantiates a tour object and takes as arguments the data \(X\),
+the tour method, e.g. \texttt{guided\_tour()}, and the starting basis.
+Once initialised, a new target plane is chosen, and a series of steps
+along a geodesic path from starting to target plane are generated by
 interpolation.
@@
 
-D3.js (Data-Driven Documents) (REF) is a JavaScript library for
-manipulating documents based on data. The advantages of D3 are similar
-to those provided by Shiny: namely, an industry standard with rich array
-of powerful, easy to use methods and widgets that can be displayed on a
-wide variety of devices, with a large user base. D3 works on data
-objects in the JavaScript Object Notation (JSON) format, which are then
-parsed and used to display customisable data visualisations.
+D3.js (Data-Driven Documents) (Bostock, Ogievetsky, and Heer 2011) is a
+JavaScript library for manipulating documents based on data. The
+advantages of D3 are similar to those provided by Shiny: namely, an
+industry standard with rich array of powerful, easy to use methods and
+widgets that can be displayed on a wide variety of devices, with a large
+user base. D3 works on data objects in the JavaScript Object Notation
+(JSON) format, which are then parsed and used to display customisable
+data visualisations.
 
@@
 \texttt{observe()} environment, which repeatedly runs the code until a
-reactive variable is changed, at which point it is is ``invalidated''
-and then re-started. The projections are calculated using the following
-code block:
+reactive variable is changed, at which point it is ``invalidated'' and
+then re-started. The projections are calculated using the following code
+block:
 
@@
 Monash University\\
-Department of Physics\\
+School of Physics and Astronomy\\
 }

### 2018-03-07 paper-tourrd3.tex
small changes

@@
 be chosen between one and the number of variables in the data. The
-display, though, has no capacity for interaction. The viewer can watch
+display, however, has no capacity for interaction. The viewer can watch
 the tour like a movie, but not pause it and restart, or change tour
@@
 These interactive controls were provided with the tourrGui package
-(Huang, Cook, and Wickham 2012), with was programmed with the RGtk2
+(Huang, Cook, and Wickham 2012), which was developed with the RGtk2
 package (Lawrence and {Temple Lang} 2010). This is not the toolkit of
-choice today, and has been superceded with primarily web-capable tools,
+choice today, and has been superseded by primarily web-capable tools,
 like Shiny (Chang et al. 2017). To display dynamic graphics though, is
-not straight-forward. This paper explains how to use D3 (Bostock,
+not straightforward. This paper explains how to use D3 (Bostock,
 Ogievetsky, and Heer 2011) as the display engine in a Shiny GUI, using
@@
 client play tag to keep up with each other, and the display cannot
-handle many observations. Noticeable slow down was oberved with 2000
-points, the main reason being the rendering time required for the large
+handle many observations. Noticeable slowdown was observed with 2000
+points, mainly due to the rendering time required for the large
 number of SVG circle elements. The situation can be improved when using

### 2018-03-15 paper-tourrd3.tex
submitted paper

--- 2018-03-02 paper-tourrd3.tex
@@
 
-The tour algorithm (Cook et al. 1995; Cook et al. 2007) is a way of
+The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995};
+Cook et al. \protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of
 systematically generating and displaying projections of high-dimensional
@@
 judged interesting according to some criterion or index function. The
-tourr package (Wickham et al. 2011) provides the computing and display
-in R (R Core Team 2018; Ihaka and Gentleman 1996) to make several types
-of tours: grand, guided, little and local. The projection dimension can
-be chosen between one and the number of variables in the data. The
-display, though, has no capacity for interaction. The viewer can watch
-the tour like a movie, but not pause it and restart, or change tour
-type, or number of variables.
+tourr package (Wickham et al. \protect\hyperlink{ref-tourr}{2011})
+provides the computing and display in R (R Core Team
+\protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
+\protect\hyperlink{ref-ihaka:1996}{1996}) to make several types of
+tours: grand, guided, little and local. The projection dimension can be
+chosen between one and the number of variables in the data. The display,
+though, has no capacity for interaction. The viewer can watch the tour
+like a movie, but not pause it and restart, or change tour type, or
+number of variables.
 
 These interactive controls were provided with the tourrGui package
-(Huang, Cook, and Wickham 2012), with was programmed with the RGtk2
-package (Lawrence and {Temple Lang} 2010). This is not the toolkit of
-choice today, and has been superceded with primarily web-capable tools,
-like Shiny (Chang et al. 2017). To display dynamic graphics though, is
-not straight-forward. This paper explains how to use D3 (Bostock,
-Ogievetsky, and Heer 2011) as the display engine in a Shiny GUI, using
-custom message passing between server and client.
+(Huang, Cook, and Wickham \protect\hyperlink{ref-tourrGui}{2012}), with
+was programmed with the RGtk2 package (Lawrence and Temple Lang
+\protect\hyperlink{ref-RGtk2}{2010}). This is not the toolkit of choice
+today, and has been superceded with primarily web-capable tools, like
+Shiny (Chang et al. \protect\hyperlink{ref-shiny}{2017}). To display
+dynamic graphics though, is not straight-forward. This paper explains
+how to use D3 (Bostock, Ogievetsky, and Heer
+\protect\hyperlink{ref-D3}{2011}) as the display engine in a Shiny
+graphical user interface (GUI), using custom message passing between
+server and client.
 
@@
 
-The \pkg{tourr} package (Wickham et al. 2011) is an R implementation of
-the tour algorithms discussed in Cook et al. (2007). It includes methods
-for geodesic interpolation and basis generation, as well as an
+The \pkg{tourr} package (Wickham et al.
+\protect\hyperlink{ref-tourr}{2011}) is an R implementation of the tour
+algorithms discussed in Cook et al.
+(\protect\hyperlink{ref-gt_pp_mc}{2007}). It includes methods for
+geodesic interpolation and basis generation, as well as an
 implementation of the simulated annealing algorithm to optimise
@@
 \begin{Sinput}
+library(tourr)
 # quartz() # to display on a Mac
 # X11() # For windows
-# The Rstudio graphics device cannot be used
-animate_dist(flea[, 1:6], center = FALSE)
+# The Rstudio graphics device is not advised
+animate_dist(flea[, 1:6], center = TRUE)
 \end{Sinput}
@@
 
-generating the animation in Figure \ref{tour}.
+generating the animation, with some stills shown in Figure \ref{tour}.
 
 \begin{figure}[ht]
-EMBED MOVIE HERE
+\centerline{\includegraphics[width=5cm]{figures/tour1.png}\includegraphics[width=5cm]{figures/tour2.png}\includegraphics[width=5cm]{figures/tour3.png}}
+\caption{Three projections from a 1D tour of 6D data, displayed as a density. Full video can be seen at https://vimeo.com/255466661.}
+\label{tour}
 \end{figure}
@@
 
-\subsection{Connecting the tour projections to D3 display using
-\texttt{sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}
+\subsection{\texorpdfstring{Connecting the tour projections to D3
+display using
+\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}
 
-D3.js (Data-Driven Documents) (Bostock, Ogievetsky, and Heer 2011) is a
-JavaScript library for manipulating documents based on data. The
-advantages of D3 are similar to those provided by Shiny: namely, an
-industry standard with rich array of powerful, easy to use methods and
-widgets that can be displayed on a wide variety of devices, with a large
-user base. D3 works on data objects in the JavaScript Object Notation
-(JSON) format, which are then parsed and used to display customisable
-data visualisations.
+D3.js (Data-Driven Documents) (Bostock, Ogievetsky, and Heer
+\protect\hyperlink{ref-D3}{2011}) is a JavaScript library for
+manipulating documents based on data. The advantages of D3 are similar
+to those provided by Shiny: namely, an industry standard with rich array
+of powerful, easy to use methods and widgets that can be displayed on a
+wide variety of devices, with a large user base. D3 works on data
+objects in the JavaScript Object Notation (JSON) format, which are then
+parsed and used to display customisable data visualisations.
 
@@
 circles in D3 language) and providing SVG objects for the web browser to
-render.
+render. Figure \ref{tourrD3} shows the new GUI.
+
+\begin{figure*}[ht]
+\centerline{\includegraphics[width=15cm]{figures/TourrD3.png}}
+\caption{Shiny GUI for the tour, with D3 as the display engine. GUI provides controls to select tour type, change speed, restart, and select variables to include.}
+\label{tourrD3}
+\end{figure*}
 
@@
 The D3 canvas makes for smooth drawing and re-drawing of the data
-projections. Adding a graphical user interface (GUI) around the display
-is straightforward with the Shiny package, e.g.~control elements such as
-stop/go, increase/decrease speed, change tour type, add/remove variables
-from the mix.
+projections. Adding a GUI around the display is straightforward with the
+Shiny package, e.g.~control elements such as stop/go, increase/decrease
+speed, change tour type, add/remove variables from the mix.
 
@@
 
-\subsection{Code}\label{code}
-
-The code is available at {[}\url{https://github.com/makipp/}
-TourrGuiD3{]}, and the source material for this paper is available at
-{[}\url{https://github.com/dicook/paper-tourrd3}{]}.
+Another disadvantage is that the displays needs to be coded anew. D3
+provides mostly primitives, and example code, to make scatterplots, and
+contours, but the data displays all need to be coded again.
 
@@
 
-The \texttt{sendCustomMessage} function provides a way to share a tour
-path with the D3 renderer, and embed it in a Shiny GUI providing
-controls such as stop/go, increase/decrease speed, change tour type,
-add/remove variables. However, the approach can result in inconsistent
-speed which appears as jerkiness in the motion, and it isn't suitable
-for large numbers of observations.
+The custom message tools from Shiny provide a way to share a tour path
+with the D3 renderer, and embed it in a Shiny GUI providing controls
+such as stop/go, increase/decrease speed, change tour type, add/remove
+variables. However, the approach doesn't provide the smooth motion that
+is needed for easy display of projections, and is slow for large numbers
+of observations.
+
+\subsection{Code}\label{code}
+
+The code is available at \url{https://github.com/makipp/TourrGuiD3}, and
+the source material for this paper is available at
+\url{https://github.com/dicook/paper-tourrd3}.
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+Thanks to Yihui Xie for pointing out the custom message tools.
 
@@
 
-\hyperdef{}{refs}{}
-
+\hypertarget{refs}{}
+\hypertarget{ref-D3}{}
 Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
@@
 
+\hypertarget{ref-shiny}{}
 Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
@@
 
+\hypertarget{ref-gt_pp}{}
 Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
@@
 
+\hypertarget{ref-gt_pp_mc}{}
 Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
@@
 
+\hypertarget{ref-tourrGui}{}
 Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
-GWidgets GUI for the Tour to Explore High-Dimensional Data Using
+gWidgets Gui for the Tour to Explore High-Dimensional Data Using
 Low-Dimensional Projections.'' \emph{Journal of Statistical Software} 49
@@
 
+\hypertarget{ref-ihaka:1996}{}
 Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
@@
 
-Lawrence, Michael, and Duncan {Temple Lang}. 2010. ``RGtk2: A Graphical
+\hypertarget{ref-RGtk2}{}
+Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
 User Interface Toolkit for R.'' \emph{Journal of Statistical Software}
@@
 
+\hypertarget{ref-R}{}
 R Core Team. 2018. \emph{R: A Language and Environment for Statistical
@@
 
+\hypertarget{ref-tourr}{}
 Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.

### 2018-03-15 submission/paper-tourrd3.Rmd
submitted paper

--- 2018-03-02 paper-tourrd3.Rmd
@@
 To display dynamic graphics though, is not straight-forward.
-This paper explains how to use D3 [@D3] as the display engine in a Shiny GUI, using custom message passing between server and client.
+This paper explains how to use D3 [@D3] as the display engine in a Shiny graphical user interface (GUI), using custom message passing between server and client.
 
@@
 ```{r eval=FALSE, echo=TRUE}
+library(tourr)
 # quartz() # to display on a Mac
 # X11() # For windows
-# The Rstudio graphics device cannot be used
-animate_dist(flea[, 1:6], center = FALSE)
+# The Rstudio graphics device is not advised
+animate_dist(flea[, 1:6], center = TRUE)
 ```
 
-generating the animation in Figure \ref{tour}.
+generating the animation, with some stills shown in Figure \ref{tour}.
 
 \begin{figure}[ht]
-EMBED MOVIE HERE
+\centerline{\includegraphics[width=5cm]{figures/tour1.png}\includegraphics[width=5cm]{figures/tour2.png}\includegraphics[width=5cm]{figures/tour3.png}}
+\caption{Three projections from a 1D tour of 6D data, displayed as a density. Full video can be seen at https://vimeo.com/255466661.}
+\label{tour}
 \end{figure}
@@
 It does this by drawing and re-drawing a scatterplot with dots (or circles in D3 language) and providing SVG objects for the web browser to render.
+Figure \ref{tourrD3} shows the new GUI.
+
+\begin{figure*}[ht]
+\centerline{\includegraphics[width=15cm]{figures/TourrD3.png}}
+\caption{Shiny GUI for the tour, with D3 as the display engine. GUI provides controls to select tour type, change speed, restart, and select variables to include.}
+\label{tourrD3}
+\end{figure*}
 
@@
 The D3 canvas makes for smooth drawing and re-drawing of the data projections.
-Adding a graphical user interface (GUI) around the display is straightforward with the Shiny package, e.g.
+Adding a GUI around the display is straightforward with the Shiny package, e.g.
 control elements such as stop/go, increase/decrease speed, change tour type, add/remove variables from the mix.
@@
 
-## Code
-
-The code is available at [https://github.com/makipp/ TourrGuiD3], and the source material for this paper is available at [https://github.com/dicook/paper-tourrd3].
+Another disadvantage is that the displays needs to be coded anew. D3 provides mostly primitives, and example code, to make scatterplots, and contours, but the data displays all need to be coded again.
 
@@
 
-The `sendCustomMessage` function provides a way to share a tour path with the D3 renderer, and embed it in a Shiny GUI providing controls such as stop/go, increase/decrease speed, change tour type, add/remove variables.
-However, the approach can result in inconsistent speed which appears as jerkiness in the motion, and it isn't suitable for large numbers of observations.
+The custom message tools from Shiny provide a way to share a tour path with the D3 renderer, and embed it in a Shiny GUI providing controls such as stop/go, increase/decrease speed, change tour type, add/remove variables.
+However, the approach doesn't provide the smooth motion that is needed for easy display of projections, and is slow for large numbers of observations.
+
+## Code
+
+The code is available at [https://github.com/makipp/TourrGuiD3](https://github.com/makipp/TourrGuiD3), and the source material for this paper is available at [https://github.com/dicook/paper-tourrd3](https://github.com/dicook/paper-tourrd3).
+
+## Acknowledgements
+
+Thanks to Yihui Xie for pointing out the custom message tools.
 

### 2019-01-21 paper-tourrd3.Rmd
Some more details on technical aspects.

@@
 
+To set up the interface for the app, we need to load the relevant scripts into the Shiny app and assign a section for the resulting plots.
+This is done when setting up the user interface.
+We import D3 and our plotting code via the `tags$script` (for web links) and `includeScript` (for reading from a full path).
+We use `tags$div` to assign an id for the output section that can be accessed in the D3 code.
+
+```
+tags$script(src = "https://d3js.org/d3.v4.min.js"),
+includeScript(system.file("js/d3anim.js", package = "tourrGUID3")),
+tags$div(id = "d3_output")
+```
+
 The data format expected by D3 is in JSON format, which combines two basic programming paradigms: a collection of name/value pairs, and an ordered list of values.
@@
 Every time a new projection has been calculated with the tour path, the resulting matrix needs to be converted to JSON and sent to D3.
+Using a named list we can send multiple JSON datasets to D3, e.g.
+to draw both the data points (stored in dataframe d) and the projection axes (stored in dataframe a).
+Converting dataframes will pass the column names to JSON.
 The code to send the D3 data looks like this:
@@
 ```
-session$sendCustomMessage(type = "data", message = toJSON(j))
+session$sendCustomMessage(type = "data", message = list(d = toJSON(d), a = toJSON(a)))
 ```
@@
 It converts the matrix of projected data points to JSON format, and sends it to JavaScript with the id data.
-When parsed in D3 by its `data()` method, it is converted back into a logical 2D array where the columns are queried first, then the rows.
-If column names are included in the JSON, the column indices are strings; otherwise they are integers starting from 1.
+The list entries of the "message" can parsed in D3 by its `data()` method, e.g.
+`data(message.d)` to access the projected data points, and we can access each column through the column names assigned in the original dataframe, and loop over all rows for rendering.
 All of the code required to render the scatterplots and legends, along with colours, is JavaScript code in the file `d3anim.js`. In particular, the data from R is handled with the following code:
@@
 
-The projections are calculated using the tour object in an `observe()` environment, which repeatedly runs the code until a reactive variable is changed, at which point it is "invalidated" and then re-started.
-The projections are calculated using the following code
-block:
+The projections are calculated using the tour object in an `observe()` environment, which re-executes the code whenever it is invalidated.
+The invalidation is either by a change in reactive value inside the code block, or we can schedule a re-execution by explicitly invalidating the observer after a selected interval.
+The projections are calculated using the following code block:
 
@@
 
+## Troubleshooting
+
+Fixing bugs in the JavaScript code can be cumbersome, as R and Shiny will not report any errors.
+Tracing JavaScript errors can be done when using the JavaScript console in the web browser.
+For example, in Google Chrome the console can be accessed via the "Developer Tools" option found under "Moore Tools" in the control menu.
+Typical errors that we encountered were version dependent syntax in D3, e.g.
+for axis definitions or scaling.
+
 ## Pros and cons

### 2019-01-21 paper-tourrd3.tex
Some more details on technical aspects.

@@
 RGtk2 package. This work explores using custom messages to pass data
-from R to D3 for viewing, using the Shiny framework.
+from R to D3 for viewing, using the Shiny framework.\\
 }
 
-\subsection{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
-The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995};
-Cook et al. \protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of
-systematically generating and displaying projections of high-dimensional
-spaces in order for the viewer to examine the multivariate distribution
-of data. It can do this either randomly, or by picking projections
-judged interesting according to some criterion or index function. The
-tourr package (Wickham et al. \protect\hyperlink{ref-tourr}{2011})
-provides the computing and display in R (R Core Team
+The tour algorithm (Cook et al. \protect\hyperlink{ref-gt_pp}{1995},
+\protect\hyperlink{ref-gt_pp_mc}{2007}) is a way of systematically
+generating and displaying projections of high-dimensional spaces in
+order for the viewer to examine the multivariate distribution of data.
+It can do this either randomly, or by picking projections judged
+interesting according to some criterion or index function. The tourr
+package (Wickham et al. \protect\hyperlink{ref-tourr}{2011}) provides
+the computing and display in R (R Core Team
 \protect\hyperlink{ref-R}{2018}; Ihaka and Gentleman
@@
 
+\hypertarget{creating-a-tour-with-the-tourr-package}{%
 \subsection{Creating a tour, with the tourr
-package}\label{creating-a-tour-with-the-tourr-package}
+package}\label{creating-a-tour-with-the-tourr-package}}
 
@@
 which instantiates a tour object and takes as arguments the data \(X\),
-the tour method, e.g. \texttt{guided\_tour()}, and the starting basis.
+the tour method, e.g.~\texttt{guided\_tour()}, and the starting basis.
 Once initialised, a new target plane is chosen, and a series of steps
@@
 
+\hypertarget{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}{%
 \subsection{\texorpdfstring{Connecting the tour projections to D3
 display using
-\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}
+\texttt{sendCustomMessage}}{Connecting the tour projections to D3 display using sendCustomMessage}}\label{connecting-the-tour-projections-to-d3-display-using-sendcustommessage}}
 
@@
 
+To set up the interface for the app, we need to load the relevant
+scripts into the Shiny app and assign a section for the resulting plots.
+This is done when setting up the user interface. We import D3 and our
+plotting code via the \texttt{tags\$script} (for web links) and
+\texttt{includeScript} (for reading from a full path). We use
+\texttt{tags\$div} to assign an id for the output section that can be
+accessed in the D3 code.
+
+\begin{verbatim}
+tags$script(src = "https://d3js.org/d3.v4.min.js"),
+includeScript(system.file("js/d3anim.js", package = "tourrGUID3")),
+tags$div(id = "d3_output")
+\end{verbatim}
+
 The data format expected by D3 is in JSON format, which combines two
@@
 with the tour path, the resulting matrix needs to be converted to JSON
-and sent to D3. The code to send the D3 data looks like this:
+and sent to D3. Using a named list we can send multiple JSON datasets to
+D3, e.g.~to draw both the data points (stored in dataframe d) and the
+projection axes (stored in dataframe a). Converting dataframes will pass
+the column names to JSON. The code to send the D3 data looks like this:
 
 \begin{verbatim}
-session$sendCustomMessage(type = "data", message = toJSON(j))
+session$sendCustomMessage(type = "data", message = list(d = toJSON(d), a = toJSON(a)))
 \end{verbatim}
@@
 file. It converts the matrix of projected data points to JSON format,
-and sends it to JavaScript with the id data. When parsed in D3 by its
-\texttt{data()} method, it is converted back into a logical 2D array
-where the columns are queried first, then the rows. If column names are
-included in the JSON, the column indices are strings; otherwise they are
-integers starting from 1. All of the code required to render the
-scatterplots and legends, along with colours, is JavaScript code in the
-file \texttt{d3anim.js}. In particular, the data from R is handled with
-the following code:
+and sends it to JavaScript with the id data. The list entries of the
+``message'' can parsed in D3 by its \texttt{data()} method,
+e.g.~\texttt{data(message.d)} to access the projected data points, and
+we can access each column through the column names assigned in the
+original dataframe, and loop over all rows for rendering. All of the
+code required to render the scatterplots and legends, along with
+colours, is JavaScript code in the file \texttt{d3anim.js}. In
+particular, the data from R is handled with the following code:
 
@@
 
-\subsection{Getting projections}\label{getting-projections}
+\hypertarget{getting-projections}{%
+\subsection{Getting projections}\label{getting-projections}}
 
@@
 The projections are calculated using the tour object in an
-\texttt{observe()} environment, which repeatedly runs the code until a
-reactive variable is changed, at which point it is ``invalidated'' and
-then re-started. The projections are calculated using the following code
-block:
+\texttt{observe()} environment, which re-executes the code whenever it
+is invalidated. The invalidation is either by a change in reactive value
+inside the code block, or we can schedule a re-execution by explicitly
+invalidating the observer after a selected interval. The projections are
+calculated using the following code block:
 
@@
 
-\subsection{Pros and cons}\label{pros-and-cons}
+\hypertarget{troubleshooting}{%
+\subsection{Troubleshooting}\label{troubleshooting}}
+
+Fixing bugs in the JavaScript code can be cumbersome, as R and Shiny
+will not report any errors. Tracing JavaScript errors can be done when
+using the JavaScript console in the web browser. For example, in Google
+Chrome the console can be accessed via the ``Developer Tools'' option
+found under ``Moore Tools'' in the control menu. Typical errors that we
+encountered were version dependent syntax in D3, e.g.~for axis
+definitions or scaling.
+
+\hypertarget{pros-and-cons}{%
+\subsection{Pros and cons}\label{pros-and-cons}}
 
@@
 
-\subsection{Summary}\label{summary}
+\hypertarget{summary}{%
+\subsection{Summary}\label{summary}}
 
@@
 
-\subsection{Code}\label{code}
+\hypertarget{code}{%
+\subsection{Code}\label{code}}
 
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
 \hypertarget{refs}{}
-\hypertarget{ref-D3}{}
+\leavevmode\hypertarget{ref-D3}{}%
 Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. ``D3:
@@
 
-\hypertarget{ref-shiny}{}
+\leavevmode\hypertarget{ref-shiny}{}%
 Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan
@@
 
-\hypertarget{ref-gt_pp}{}
+\leavevmode\hypertarget{ref-gt_pp}{}%
 Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
@@
 
-\hypertarget{ref-gt_pp_mc}{}
+\leavevmode\hypertarget{ref-gt_pp_mc}{}%
 Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2007.
@@
 
-\hypertarget{ref-tourrGui}{}
+\leavevmode\hypertarget{ref-tourrGui}{}%
 Huang, Bei, Dianne Cook, and Hadley Wickham. 2012. ``TourrGui: A
@@
 
-\hypertarget{ref-ihaka:1996}{}
+\leavevmode\hypertarget{ref-ihaka:1996}{}%
 Ihaka, Ross, and Robert Gentleman. 1996. ``R: A Language for Data
@@
 
-\hypertarget{ref-RGtk2}{}
+\leavevmode\hypertarget{ref-RGtk2}{}%
 Lawrence, Michael, and Duncan Temple Lang. 2010. ``RGtk2: A Graphical
@@
 
-\hypertarget{ref-R}{}
+\leavevmode\hypertarget{ref-R}{}%
 R Core Team. 2018. \emph{R: A Language and Environment for Statistical
@@
 
-\hypertarget{ref-tourr}{}
+\leavevmode\hypertarget{ref-tourr}{}%
 Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
@@
 Projections.'' \emph{Journal of Statistical Software} 40 (2): 1--18.
+
+\bibliography{references.bib}
 

### 2019-03-03 paper-tourrd3.Rmd
revisions to make it easier to understand

@@
     email:  mkipp271@gmail.com
+  - name: Ursula Laa
+    affiliation: Monash University
+    address:
+    - School of Physics and Astronomy
+    email:  ursula.laa@monash.edu
   - name: Dianne Cook
@@
     email:  dicook@monash.edu 
-  - name: Ursula Laa
-    affiliation: Monash University
-    address:
-    - School of Physics and Astronomy
-    email:  ursula.laa@monash.edu
 bibliography: references.bib
@@
 This work explores using custom messages to pass data from R to D3  for viewing, using the Shiny framework.
+This is an approach that can be generally used for creating all sorts of interactive graphics.
 preamble: >
@@
 library(RefManageR)
-knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, kfigr.link=TRUE, kfigr.prefix=TRUE, cache=TRUE, fig.env = TRUE, fig.cap=TRUE, fig.height=3)
+knitr::opts_chunk$set(echo=FALSE, 
+                      message=FALSE, 
+                      warning=FALSE, 
+                      kfigr.link=TRUE, 
+                      kfigr.prefix=TRUE, 
+                      cache=TRUE, 
+                      fig.env = TRUE, 
+                      fig.cap=TRUE, 
+                      fig.height=3)
 
@@
 ## Introduction
+
+Dd you know you can run *any javascript you like* in a Shiny application and you can pass *whatever you want including JSON* back and forth?
+This massively widens the scope of what you can do with Shiny, and generating a tour of multivariate data with this approach is a really good example of what is possible.
 
@@
 It includes methods for geodesic interpolation and basis generation, as well as an implementation of the simulated annealing algorithm to optimise projection pursuit indices for the guided tour.
-The tour can be displayed directly in the R graphics device, for example:
+The tour can be displayed directly in the R graphics device, for example, the code below generates a 1D density tour.
+Figure \ref{tour} shows snapshots.
 
-```{r eval=FALSE, echo=TRUE}
+```{r eval=FALSE, echo=TRUE, tidy=TRUE}
 library(tourr)
-# quartz() # to display on a Mac
-# X11() # For windows
+# quartz() # to display on a Mac;
+# X11() # For windows;
 # The Rstudio graphics device is not advised
@@
 ```
-
-generating the animation, with some stills shown in Figure \ref{tour}.
 
@@
 
+## Try it
+
+You can try the app yourself using this code:
+
+```{r eval=FALSE, echo=TRUE, tidy=TRUE}
+devtools::install_github("uschiLaa/tourrGUID3")
+library(tourrGUID3)
+launchApp(system.file("extdata", "geozoo.csv", package = "tourrGUID3"))
+```
+
 ## Troubleshooting

### 2019-03-03 paper-tourrd3.tex
revisions to make it easier to understand

@@
 with tours}
-\author{by Michael Kipp, Dianne Cook, Ursula Laa}
+\author{by Michael Kipp, Ursula Laa, Dianne Cook}
 
@@
 RGtk2 package. This work explores using custom messages to pass data
-from R to D3 for viewing, using the Shiny framework.\\
+from R to D3 for viewing, using the Shiny framework. This is an approach
+that can be generally used for creating all sorts of interactive
+graphics.\\
 }
@@
 \subsection{Introduction}\label{introduction}}
+
+Dd you know you can run \emph{any javascript you like} in a Shiny
+application and you can pass \emph{whatever you want including JSON}
+back and forth? This massively widens the scope of what you can do with
+Shiny, and generating a tour of multivariate data with this approach is
+a really good example of what is possible.
 
@@
 projection pursuit indices for the guided tour. The tour can be
-displayed directly in the R graphics device, for example:
+displayed directly in the R graphics device, for example, the code below
+generates a 1D density tour. Figure \ref{tour} shows snapshots.
 
@@
 library(tourr)
-# quartz() # to display on a Mac
-# X11() # For windows
-# The Rstudio graphics device is not advised
+# quartz() # to display on a Mac; X11() # For windows; The Rstudio graphics
+# device is not advised
 animate_dist(flea[, 1:6], center = TRUE)
@@
 \end{Schunk}
-
-generating the animation, with some stills shown in Figure \ref{tour}.
 
@@
 which instantiates a tour object and takes as arguments the data \(X\),
-the tour method, e.g.~\texttt{guided\_tour()}, and the starting basis.
+the tour method, e.g. \texttt{guided\_tour()}, and the starting basis.
 Once initialised, a new target plane is chosen, and a series of steps
@@
 and sends it to JavaScript with the id data. The list entries of the
-``message'' can parsed in D3 by its \texttt{data()} method,
-e.g.~\texttt{data(message.d)} to access the projected data points, and
-we can access each column through the column names assigned in the
-original dataframe, and loop over all rows for rendering. All of the
-code required to render the scatterplots and legends, along with
-colours, is JavaScript code in the file \texttt{d3anim.js}. In
-particular, the data from R is handled with the following code:
+``message'' can parsed in D3 by its \texttt{data()} method, e.g.
+\texttt{data(message.d)} to access the projected data points, and we can
+access each column through the column names assigned in the original
+dataframe, and loop over all rows for rendering. All of the code
+required to render the scatterplots and legends, along with colours, is
+JavaScript code in the file \texttt{d3anim.js}. In particular, the data
+from R is handled with the following code:
 
@@
 \end{verbatim}
+
+\hypertarget{try-it}{%
+\subsection{Try it}\label{try-it}}
+
+You can try the app yourself using this code:
+
+\begin{Schunk}
+\begin{Sinput}
+devtools::install_github("uschiLaa/tourrGUID3")
+library(tourrGUID3)
+launchApp(system.file("extdata", "geozoo.csv", package = "tourrGUID3"))
+\end{Sinput}
+\end{Schunk}
 
@@
 \address{%
+Ursula Laa\\
+Monash University\\
+School of Physics and Astronomy\\
+}
+\href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}
+
+\address{%
 Dianne Cook\\
@@
 \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}
-
-\address{%
-Ursula Laa\\
-Monash University\\
-School of Physics and Astronomy\\
-}
-\href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}

### 2019-03-03 review_rebuttal.Rmd
added details to rebuttal

---
title: "Review rebuttal"
author: "Michael Kipp, Ursula Laa, Di Cook"
date: "01/03/2019"
output: html_document
---

*We are pleased to accept your paper "Connecting R with D3 for dynamic graphics, to explore multivariate data with tours" subject to the major revisions described below.*

*Though the reviewer was critical, they thought there were enough interesting parts to form a paper around, though they expect that there will be some fleshing out, as described below:*

*> Overall recommendation: Accept with major revisions.*

*I'm very underwhelmed by this paper, to be honest. I think it's a bit of a tricky one to get excited about because all the Shiny experts will be reading thinking "well of course I knew you could do that" whereas all the people less familiar with Shiny won't really understand the approach. Other than that there's nothing fundamentally wrong with the approach, it's a sensible solution to a real problem and they've thoughtfully evaluated the pros and cons at the end.*

**Thanks, this summary is appreciated**

*> Overview: Is the article important to R the community? Is the approach sensible? Are technology and methods up-to-date?*

*Yes, it's sort of minimally important, yes. I think to make it feel more relevant it would be better to reframe as saying "hey, did you know you can run ANY JAVASCRIPT YOU LIKE in a Shiny application and you can pass WHATEVER YOU WANT INCLUDING JSON back and forth, this massively widens the scope of what you can do with Shiny, here's a really good example of what we did". It seems a bit churlish, I suppose, as well, but having said the tourrGui package is out of date it makes you think maybe they should turn theirs into a package*

**Additional sentences have been added to the introduction and abstract to indicate the general nature of the basic tool.**

*> Article: Is anything unclear? Is anything missing? Are the examples helpful? Is there sufficient background, including alternative approaches in R?**

*For the novices, it would be nice for them to include the method to run their code straight from the terminal:*

```
library(shiny)
runGitHub("makipp/TourrGuiD3")
```

**Thank you, a shiny app has been created and these instructions will run it.**

```
devtools::install_github("uschiLaa/tourrGUID3")
library(tourrGUID3)
launchApp(system.file("extdata", "geozoo.csv", package = "tourrGUID3"))
```

*I really think they need to make the paper more specialised to appeal to a Shiny audience, telling people who know Shiny stuff they don't already know, or they need to make it more general, filling in a few blanks for people who don't know Shiny. For example, they skip straight over the invalidateLater() function, they could at least tell us what it's doing.*

**More explanation has been added**

*They also say "The projections are calculated using the tour object in an observe() environment, which repeatedly runs the code until a reactive variable is changed, at which point it is “invalidated” and then re-started". I think this is a bit confusing. In fact the code ONLY reruns when it is invalidated (whether by invalidateLater() or a reactive value), that's how observe() works. I'm sure that's what they mean, really, but it isn't expressed very well.*

**Change made**

*I'm not over the moon about the dependency on ggvis, it's not that important since this is hardly production code, but this package is officially dormant and I think it's best avoided where possible, to save readers from going down a blind alley with it.*

**Dependency on ggvis removed, it was unnecessary, and yes, was simply left over from an older version used in testing.**

*The last thing is I would like them just to give one sentence to explain how the D3 actually ends up on the page. Really basic stuff for your Shiny savvy readers, but I don't think it is of much interest to them anyway because it's all basic stuff for the Shiny savvy.*

**check this**

*> Package: Could the functions presented be improved? Some ideas at http://bit.ly/1aYwkp3.*

*I think the code is fine. For me this paper could be improved by explicitly broadening the scope and hand holding a bit for people who are not totally au fait with Shiny.*

**check this**

*We would appreciate a revised version and point-by-point response to the reviewers comments within 2 months. Remember, that when responding to the reviewer's comments, your job is to persuade me, the editor, that either you've dealt with the issue, or that it's not relevant. To this end, please produce a single document that includes all the reviewers comments mingled with your responses. I particularly recommend the strategy described at http://matt.might.net/articles/peer-review-rebuttals/.*

### 2019-03-03 paper-tourrd3.Rmd
Some more details about D3 interface.

@@
 
-Dd you know you can run *any javascript you like* in a Shiny application and you can pass *whatever you want including JSON* back and forth?
+Did you know you can run *any javascript you like* in a Shiny application and you can pass *whatever you want including JSON* back and forth?
 This massively widens the scope of what you can do with Shiny, and generating a tour of multivariate data with this approach is a really good example of what is possible.
@@
 
+On the D3 side we can access the id defined in Shiny, and for example assign it to a scalable vector graphics (svg) object to be filled in D3 and rendered onto the Shiny app.
+
+```
+var svg = d3.select("#d3_output")
+    .append("svg")
+    .attr("width", w)
+    .attr("height", h);
+```
+
 The data format expected by D3 is in JSON format, which combines two basic programming paradigms: a collection of name/value pairs, and an ordered list of values.
@@
               input$guidedIndex,
-              c(rv$class[[1]]), 
-              input$scagType), b)
+              c(rv$class[[1]])), b)
 })
@@
 The projections are calculated using the tour object in an `observe()` environment, which re-executes the code whenever it is invalidated.
-The invalidation is either by a change in reactive value inside the code block, or we can schedule a re-execution by explicitly invalidating the observer after a selected interval.
+The invalidation is either by a change in reactive value inside the code block, or we can schedule a re-execution by explicitly invalidating the observer after a selected interval using `invalidateLater()`.
 The projections are calculated using the following code block:
@@
 observe({
+      if (length(rv$mat[1, ]) < 3) {
+        session$sendCustomMessage(type = "debug",
+        message = "Error: Need >2 variables.")
+      }
+      aps <- rv$aps
       tour <- rv$tour
-  aps <- rv$aps
-  step <- tour(aps / fps)
-  if (!is.null(step)) {
+      step <- rv$tour(aps / fps)
       invalidateLater(1000 / fps)
@@
       session$sendCustomMessage(type = "data",
-      message = toJSON(j))
-  }
-  else {
-    if (length(rv$mat[1, ]) < 3) {
-    session$sendCustomMessage(type = "debug", 
-      message = "Error: Need > 2 variables.")
-    } else {
-    session$sendCustomMessage(type = "debug", 
-    message = "Guided tour finished: no better bases found.")
-    }
-  }
+                                message =
+                                list(d = toJSON(data.frame(pL=rv$pLabel[,1],x=j[,2],y=j[,1],c=j[,3])),
+                                a = toJSON(data.frame(n=rv$vars,y=step$proj[,1],x=step$proj[,2]))))
 })

### 2019-03-03 paper-tourrd3.tex
Some more details about D3 interface.

@@
 
-Dd you know you can run \emph{any javascript you like} in a Shiny
+Did you know you can run \emph{any javascript you like} in a Shiny
 application and you can pass \emph{whatever you want including JSON}
@@
 library(tourr)
-# quartz() # to display on a Mac; X11() # For windows; The Rstudio graphics
-# device is not advised
+# quartz() # to display on a Mac;
+# X11() # For windows;
+# The Rstudio graphics device is not advised
 animate_dist(flea[, 1:6], center = TRUE)
@@
 
+On the D3 side we can access the id defined in Shiny, and for example
+assign it to a scalable vector graphics (svg) object to be filled in D3
+and rendered onto the Shiny app.
+
+\begin{verbatim}
+var svg = d3.select("#d3_output")
+    .append("svg")
+    .attr("width", w)
+    .attr("height", h);
+\end{verbatim}
+
 The data format expected by D3 is in JSON format, which combines two
@@
 and sends it to JavaScript with the id data. The list entries of the
-``message'' can parsed in D3 by its \texttt{data()} method, e.g.
-\texttt{data(message.d)} to access the projected data points, and we can
-access each column through the column names assigned in the original
-dataframe, and loop over all rows for rendering. All of the code
-required to render the scatterplots and legends, along with colours, is
-JavaScript code in the file \texttt{d3anim.js}. In particular, the data
-from R is handled with the following code:
+``message'' can parsed in D3 by its \texttt{data()} method,
+e.g.~\texttt{data(message.d)} to access the projected data points, and
+we can access each column through the column names assigned in the
+original dataframe, and loop over all rows for rendering. All of the
+code required to render the scatterplots and legends, along with
+colours, is JavaScript code in the file \texttt{d3anim.js}. In
+particular, the data from R is handled with the following code:
 
@@
 inside the code block, or we can schedule a re-execution by explicitly
-invalidating the observer after a selected interval. The projections are
-calculated using the following code block:
+invalidating the observer after a selected interval using
+\texttt{invalidateLater()}. The projections are calculated using the
+following code block:
 

### 2019-03-04 paper-tourrd3.Rmd
link to app code to Ursulas page

@@
 
-The code is available at [https://github.com/makipp/TourrGuiD3](https://github.com/makipp/TourrGuiD3), and the source material for this paper is available at [https://github.com/dicook/paper-tourrd3](https://github.com/dicook/paper-tourrd3).
+The code is available at [https://github.com/uschiLaa/tourrGUID3](https://github.com/uschiLaa/tourrGUID3), and the source material for this paper is available at [https://github.com/dicook/paper-tourrd3](https://github.com/dicook/paper-tourrd3).
 

### 2019-03-04 paper-tourrd3.tex
link to app code to Ursulas page

@@
 library(tourr)
-# quartz() # to display on a Mac;
-# X11() # For windows;
-# The Rstudio graphics device is not advised
+# quartz() # to display on a Mac; X11() # For windows; The Rstudio graphics
+# device is not advised
 animate_dist(flea[, 1:6], center = TRUE)
@@
 
-The code is available at \url{https://github.com/makipp/TourrGuiD3}, and
-the source material for this paper is available at
+The code is available at \url{https://github.com/uschiLaa/tourrGUID3},
+and the source material for this paper is available at
 \url{https://github.com/dicook/paper-tourrd3}.

### 2019-03-04 paper-tourrd3.Rmd
code reformatting

@@
 
-```
+```{r eval=FALSE, echo=TRUE}
 observe({
@@
     session$sendCustomMessage(type = "data",
-                                message =
-                                list(d = toJSON(data.frame(pL=rv$pLabel[,1],x=j[,2],y=j[,1],c=j[,3])),
-                                a = toJSON(data.frame(n=rv$vars,y=step$proj[,1],x=step$proj[,2]))))
+         message = list(d = toJSON(data.frame(pL=rv$pLabel[,1], x=j[,2], 
+                                              y=j[,1], c=j[,3])),
+                        a = toJSON(data.frame(n=rv$vars, y=step$proj[,1], 
+                                              x=step$proj[,2]))))
 })

### 2019-03-04 paper-tourrd3.tex
code reformatting

@@
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 observe({
@@
     session$sendCustomMessage(type = "data",
-                                message =
-                                list(d = toJSON(data.frame(pL=rv$pLabel[,1],x=j[,2],y=j[,1],c=j[,3])),
-                                a = toJSON(data.frame(n=rv$vars,y=step$proj[,1],x=step$proj[,2]))))
+         message = list(d = toJSON(data.frame(pL=rv$pLabel[,1], x=j[,2], 
+                                              y=j[,1], c=j[,3])),
+                        a = toJSON(data.frame(n=rv$vars, y=step$proj[,1], 
+                                              x=step$proj[,2]))))
 })
-\end{verbatim}
+\end{Sinput}
+\end{Schunk}
 
